---
title: "MBNI_AffymetrixRae230_ F15"
date: "Affymetrix microarray study on P14 HR/LR"

output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_knit$set(root.dir = "~/Phenotype Project/ibirt")
```

The file includes code for the re-analysis of the Sarah Clinton MBNI Affy P14 data set. The data set is RMA data, it has been background corrected, normalized and annotated. The directory structure for this analysis mirrors the structure for the others.

```{r loadData, warning=FALSE}

library(car)
library(plyr)
library(knitr)

dataDir <- "data/MBNI Affy P14/"
outDir <- "output/MBNI Affy P14/"
qcDir <- "qc/MBNI Affy P14/"

affyData <- read.csv(paste0(dataDir, "RMAExpression_customCDF.csv"))
affyAnnotation <- read.csv(paste0(dataDir, "RMAExpression_customCDFAnnotation2.csv"))

```

## Outlier Checks
Box plots and heatmaps to check for outliers.

```{r outlierChecks, fig.align='center', fig.width=7}

# box plot of the data
boxplot(affyData[, c(2:13)], cex=0.25, las=3, par(cex.axis=0.75),
        main="Log signal values per sample (1 box = all filtered probes)",
        xlab="Sample ID", ylab="Log Signal")
```

Looks fine so far. Let's check sample correlations.

```{r correlations, fig.align='center', fig.width=7}

logDataMatrix <- as.matrix(affyData[, -1])

# get the correlations
corMatrix <- cor(logDataMatrix)

# write it to file
write.csv(corMatrix, paste0(qcDir, "logdataCormatrix.csv"))

# the heatmap
heatmap(corMatrix)

affyData$X6_RN230_HC_P14_F15_L03.CEL <- NULL
```

The heatmap shows that L03 is an outlier. That id is removed from the data analysis.

## Annotation
The following code chunk annotates the data.

```{r annotate}

affyA <- join(affyAnnotation, affyData, by="ProbesetID", type="inner")

```

## Normalization
The following code chunk averages by gene symbol (there are 2 duplicates after removiing NAs) and then scales and centers the data. 

```{r normalize}

# let's remove NA's
affynoNA <- affyA[which(!is.na(affyA$GeneSymbol)),]

# Check for duplicates
sum(duplicated(affynoNA$GeneSymbol))
#[1] 2

# There are two duplicates. We average by gene symbol to remove the duplicates.
valueCols <- grep("CEL", colnames(affynoNA))
valueCols <- colnames(affynoNA)[valueCols]

affyAverage <- sapply(valueCols, function(x, frame) {
  tapply(frame[[x]], frame$GeneSymbol, mean)
}, frame = affynoNA)

# no need to set colnames and row names as they are automatically set.

# get the z-scores
affyZscore <- t(scale(t(affyAverage), scale = TRUE, center = TRUE))
affyZscore <- cbind.data.frame(GeneSymbol = row.names(affyAverage),
                                affyZscore)  #adding back row of genesymbols

```

## Cell Type Analysis

The following code matches the gene symbols to cell types and gets the average of the z-score by cell type.

```{r celltype}

cellInfo <- read.csv("data/CellTypeSpecificGenes_Master3.csv")
colnames(cellInfo)[5] <- "GeneSymbol"

affyCellType <- join(affyZscore, cellInfo,
                     by = "GeneSymbol", type = "inner")

# remove repeating cell types
noPrimaryOverlap <- lapply(levels(affyCellType$CellType_Primary),
                           function(x, celltype) {
  
  # choose all data for a particular primary cell type
  tempData <- celltype[which(celltype$CellType_Primary == x), ]
  
  # get all gene symbols within the current primary cell type
  tempGenes <- tempData$GeneSymbol
  
  # gene symbols within all toher cell types
  otherGenes <- celltype$GeneSymbol[which(celltype$CellType_Primary != x)]
  
  # now grab the rows which are not in otherGenes
  return(tempData[which(!(tempGenes %in% otherGenes)), ])
}, celltype = affyCellType)

# combine the results
noPrimaryOverlap <- do.call(rbind, noPrimaryOverlap)

# some genes are duplicated
sum(duplicated(noPrimaryOverlap$GeneSymbol))

#Removing duplicate genes
noPrimaryOverlap<-subset(noPrimaryOverlap, 
                         !duplicated(noPrimaryOverlap$GeneSymbol))

# average over cell types
avgCellType <- sapply(valueCols, function(x, frame) {
  tapply(frame[[x]], frame[["CellType_Primary"]], mean, na.rm = TRUE)
}, frame = noPrimaryOverlap)

```

## Plot the Phenotype Effects

The following code chunk makes the data frame for the cell type phenotype effects and prints the boxplots for each cell type by phenotype.

```{r phenBox, fig.align='center', fig.width=7}

# No particular order
phenotype <- factor(c("HR", "LR", "HR", "LR", "LR", "HR", "LR", "HR", "HR", "LR", "HR"))

phenEffectCellType <- cbind.data.frame(phenotype, t(avgCellType))

# draw the box plots. Print to PDF output rather than separate files.
for (i in 2:11) {
  
  boxplot(phenEffectCellType[[i]] ~ phenEffectCellType$phenotype,
          ylab = row.names(avgCellType)[i], 
          main=paste(row.names(avgCellType)[i], "vs Phenotype"),
          pch=20, col=(i+5), border=(i))
  
}


```

Now moving on to the phenotype ANOVA. Will also print this in-file and save the p-values.

```{r phenAnova, results='asis'}

cNames <- colnames(phenEffectCellType)[2:11]

anovaPhen <- lapply(phenEffectCellType[cNames], function(x, dataFrame) {
  Anova(lm(x ~ phenotype, data = dataFrame, 
           contrasts = list(phenotype=contr.sum)), 
        type = 3)
  }, dataFrame = phenEffectCellType)

# Store it in matrix form
CellTypeAnovaPvals <- matrix(0, 10, 1)
colnames(CellTypeAnovaPvals) <- c("p-value")
row.names(CellTypeAnovaPvals) <- colnames(phenEffectCellType[2:11])

for (i in 1:10) {
  CellTypeAnovaPvals[i, ] <- anovaPhen[[i]]$`Pr(>F)`[2]
}

# Save it
write.csv(CellTypeAnovaPvals, 
          paste0(outDir, "CellTypeAnovaPvals_P14Affy230.csv"))


# Now, anovaPhen has all the ANOVA stuff. Print it.
for(i in 1:length(anovaPhen)){
  print(kable(anovaPhen[[i]], caption=names(anovaPhen)[i]))
}

```

## Phenotype Effect by Gene Symbol

Now run the ANOVA for genes. The process is quite the same.

```{r phenGene}

# keep only the numeric data
affyZscore$GeneSymbol <- NULL

zscoreByGeneTranspose <- cbind.data.frame(phenotype, t(affyZscore))

# Now, loop over all genes to get the ANOVA
anovaOutput <- sapply(row.names(affyZscore), function(x, frame) {
  
  Anova(lm(frame[[x]] ~ phenotype,
           data = frame,
           contrasts = c(phenotype = contr.sum)),
        type = 3)$"Pr(>F)"[2]
  
}, frame = zscoreByGeneTranspose)

phenotypeEffects <- cbind.data.frame("GeneSymbol" = names(anovaOutput),
                                     "p-value" = anovaOutput)

phenotypeEffectsCellInfo <- join(phenotypeEffects, cellInfo,
                                 by="GeneSymbol", type="left")

# write to file
write.csv(phenotypeEffectsCellInfo,
          paste0(outDir, "PhenPvals_CellInfo_P14Affy230.csv"))

```

## T-tests for meta-analysis

The following code chunk does the t-tests for the meta-analysis and writes the results to a file.

```{r metaAnalysis}

#Join average by gene symbol log data with phenotype info
affytt <- cbind.data.frame(phenotype, t(affyAverage))

phenTtests <- sapply(row.names(affyAverage), function(x, frame) {
  
  t.test(frame[[x]] ~ phenotype, data = frame)$statistic
  
}, frame = affytt)

# write to file
phenTtests <- as.data.frame(phenTtests)

#Code adds .t after row.names so need to reannotate
row.names(phenTtests) <- row.names(affyAverage)
colnames(phenTtests) <- c("Affy P14 t-test")

write.csv(phenTtests, paste0(outDir, "affy230P14TT.csv"))

# Cell type t-tests for meta-analysis

cellTypeTtests <- sapply(colnames(phenEffectCellType)[-1], function(x, frame) {
  
  t.test(frame[[x]] ~ phenotype, data = frame)$statistic
  
}, frame = phenEffectCellType)

cellTypeTtests <- as.data.frame(cellTypeTtests)

#reannotate
row.names(cellTypeTtests)<- colnames(phenEffectCellType)[-1]
colnames(cellTypeTtests) <- c("Affy P14 t-test")

write.csv(cellTypeTtests,
          paste0(outDir, "P14Affy230CellTypeTT_forMeta.csv"))

```
